---
title: "Project_Health Insurance Analysis"
author: "Akshat Badhwar"
date: "2025-08-22"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importing the Dataset
```{r}
setwd('C:/Users/aksha/Downloads')
insurance <- read.csv('insurance.csv')
```

# Exploratory Data Analysis (EDA)

```{r}
str(insurance) # Structure of dataset (Data types and Columns)
```
```{r}
summary(insurance) # Summary statistics of all variables
```

```{r}
outliers <- boxplot.stats(insurance$charges)$out # Identifies outliers in 'Charges'
outliers
length(outliers) # Count of the number of outliers
```
```{r}
boxplot(insurance$charges, 
        main = "Charges by Smoking Status", ylab = "Charges", 
        col = "lightgreen", horizontal = TRUE)    # Visualises outliers
```

# Descriptive Statistics and Visualisation
```{r}
hist(insurance$age, main = 'Distribution of Ages', xlab = 'Age', col = 'yellow')
```
```{r}
hist(insurance$bmi, main = 'Distribution of BMI', xlab = 'BMI', col = 'orange', breaks = 20)
```
```{r}
hist(insurance$charges, main = 'Distribution of Insurance Charges', xlab = 'Charges', col = 'lightblue', breaks = 30)
```
```{r}
boxplot(charges ~ smoker, data = insurance, col=c("lightblue","pink"))
```

# Average Charges by Smoking Status
```{r}
insurance$smoker <- trimws(tolower(insurance$smoker)) # Cleans 'Smoker' variable
```
```{r}
unique(insurance$smoker) # Verifies unique values
```

```{r}
avg_charges <- tapply(insurance$charges, insurance$smoker, mean, na.rm = TRUE)
```
```{r}
names(avg_charges) <- c("Non-Smokers", "Smokers")
```

```{r}
barplot(avg_charges, 
        main = "Average Insurance Charges: Smokers vs Non-Smokers",
        col = c("skyblue", "salmon"),
        ylab = "Average Charges",
        xlab = "Smoking Status")
```

# Hypothesis Testing (Test 1: t-test)
H₀: Average charges for Smokers = Non-smokers
H₁: Average charges for Smokers ≠ Non-smokers

```{r}
smoker_charges <- insurance$charges[insurance$smoker == "yes"]
```
```{r}
nonsmoker_charges <- insurance$charges[insurance$smoker == "no"]
```
```{r}
t.test(charges ~ smoker, data = insurance)
```
p-value < 0.05 (H0 is rejected)
Average charges for Smokers is not equal to that of Non-Smokers

# Hypothesis Testing (Test 2: ANOVA Test)
H₀: Mean Premiums are equal across regions
H₁: Mean of at least one region is different

```{r}
anova_result <- aov(charges ~ region, data=insurance)
```
```{r}
summary(anova_result)
```
```{r}
TukeyHSD(anova_result) # Identifies mean of which regions are significantly different
```

p-value < 0.05 (H0 is rejected)
Mean of at least one region is different 

# Hypothesis Testing (Test 3: Chi-Square Test) 
H₀: Smoking and gender are independent (no relationship)
H₁: Smoking and gender are dependent (there is a relationship)

```{r}
smoke_gender <- table(insurance$smoker, insurance$sex)
```
```{r}
chi_result <- chisq.test(smoke_gender)
```
```{r}
chi_result
```

p-value < 0.05 (H0 is rejected)
Smoking and gender are related

# Correlation Analysis
```{r}
round(cor(insurance[c("age","bmi","children","charges")]), 2) # Shows how strongly numeric variables are related
```

# Regression Analysis and Model Diagnostics

```{r}
model <- lm(charges ~ age + bmi + children + smoker + sex + region, data=insurance)
```
```{r}
summary(model) # Checks coefficients and significance
```

# Diagnostic plots (Residuals, Normality, Leverage)
```{r}
par(mfrow=c(2,2))
```
```{r}
plot(model)
```
```{r}
par(mfrow=c(1,1))
```
